---
title: "ghrsst"
output: github_document
---

Access OPeNDAP GHRSST data from R. Extract point or bounded boxes (as raster).

## Requirements

+ [R v4.1+](https://www.r-project.org/)

Packages from CRAN.

+ [rlang](https://CRAN.R-project.org/package=rlang)
+ [dplyr](https://CRAN.R-project.org/package=httr)
+ [sf](https://CRAN.R-project.org/package=sf)
+ [stars](https://CRAN.R-project.org/package=stars)
+ [tidyr](https://CRAN.R-project.org/package=tidyr)
+ [ncdf4](https://CRAN.R-project.org/package=ncdf4)

Packages from Github

+ [xyzt](https://github.com/BigelowLab/xyzt)


## Installation

```
remotes::install_github("BigelowLab/ghrsst")
```

### Usage
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(sf)
  library(ghrsst)
  library(xyzt)
  library(stars)
})
```

#### Working with points.  

See the [xyzt](https://github.com/BigelowLab/xyzt) package for more details on the example Gulf of Maine data.

```{r}
# read in example GOM points
x <- xyzt::read_gom() |>
  dplyr::select(-time, -depth) |>
  xyzt::as_POINT()

# generate a MUR url for a given date
url <- mur_url("2018-12-18")

# open the resource
X <- ncdf4::nc_open(url)

# extract the data 
covars <- ghrsst::extract(x, X, varname = mur_vars(X))

# bind to the input
(y <- dplyr::bind_cols(x, covars))
```

#### Working with bounding boxes (from points or polygons). 

Learn more about working with [stars](https://CRAN.R-project.org/package=stars) objects in the [vignettes](https://r-spatial.github.io/stars/).

```{r}
# read in example GOM points
x <- xyzt::read_gom() |>
  dplyr::select(-time, -depth) |>
  xyzt::as_BBOX()

(covars <- ghrsst::extract(x, X, varnames = ghrsst::mur_vars(X)))
```


```{r}
# cleanup
ncdf4::nc_close(X)
```
